Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-20 ind_files/05_unification_discussion.tex
Minor rephrasing

More generally, the well-known intimate relation between It\^o SDEs and Fokker-Planck equations (see section \ref{sec_math_background}) is extremely useful, and indeed has been extensively exploited~\citep{van_kampen_stochastic_1981,oksendal_stochastic_1998, gardiner_stochastic_2009}, because some problems are much more easily attacked using probabilistic tools that operate on It\^o processes, whereas others are more amenable to study using tools from PDE theory that operate on the Fokker-Planck equations. In the infinite-dimensional case, working with the field theory through functional Fokker-Planck equations such as \eqref{disc_functional_field_eqns} allows us to use tools with an essentially `differential equation' flavor such as spectral methods (\cite{rogers_demographic_2012, rogers_modes_2015}; See also Appendix \ref{App_examples} for a general model-independent pedagogical example for studying phenotypic clustering and evolutionary branching), slow manifold (`adiabatic') approximations~\citep{parsons_dimension_2017}, and WKB style expansions~(\cite{rogers_demographic_2012,assaf_wkb_2017}; though it bears noting that WKB theory is closely related to large deviation principles that occur in the study of Markov processes and come in a distinctly `probabilistic' flavor, see~\cite{dembo_large_1998}). On the other hand, many biologically interesting questions are much easier to approach from the branching processes/SPDE side using tools that have an essentially `probabilistic' flavor such as duality~\citep{dawson_stochastic_1975,greenman_duality_2020}, infinitesimal generators~\citep{ethier_markov_1986,etheridge_mathematical_2011}, martingale theory~\citep{champagnat_unifying_2006,etheridge_mathematical_2011}, and weak/mild solutions to SPDEs~\citep{week_white_2021}. Rigorously establishing and systematically exploiting relations between functional Fokker-Planck equations of the form \eqref{disc_functional_field_eqns} and the sort of general SPDEs studied in~\cite{week_white_2021} for measure-valued branching processes could thus prove very fruitful for developing more integrative eco-evolutionary theory since it allows us to seamlessly transition between complementary views of the same object to suit the problem at hand\footnote{The physics literature has long recognized this. Physicists regularly exploit the connections between stochastic/quantum field equations and SPDEs, often studying SPDEs in \emph{classical} systems using tools and language from quantum field theory~\citep{hochberg_effective_1999,baez_quantum_2018}.}.

@@

Currently, (stochastic) field equations are primarily used by physicists working in areas such as statistical field theory and quantum field theory, and are largely attacked using ingenious and often somewhat system-specific tools that may not necessarily be mathematically well understood~(\cite{carmona_stochastic_1999}; but see \cite{bogachev_fokker-planck-kolmogorov_2015} for a rigorous treatment of some infinite-dimensional Fokker-Planck equations). Equation \eqref{disc_functional_field_eqns} opens up the study of quantitative trait dynamics in finite fluctuating populations to analysis using some of these tools - For example, a careful reading of the literature provides many hints (albeit couched in physics language) for how (quantum) field theoretic ideas such as the Fock space representation~\citep{del_razo_probabilistic_2022}, `loop diagrams/expansions'~\citep{hochberg_effective_1999,dodd_many-body_2009}, and the often closely related path integral formalism~\citep{doi_second_1976, peliti_path_1985,hochberg_effective_1999, chow_path_2015, weber_master_2017}, can all be co-opted to coax biological insights from the sort of stochastic processes modeled by \eqref{disc_functional_field_eqns}.

>>>>

More generally, the well-known intimate relation between It\^o SDEs and Fokker-Planck equations (see section \ref{sec_math_background}) is extremely useful, and indeed has been extensively exploited~\citep{van_kampen_stochastic_1981,oksendal_stochastic_1998, gardiner_stochastic_2009}, because some problems are much more easily attacked using probabilistic tools that operate on It\^o processes, whereas others are more amenable to study using tools from PDE theory that operate on the Fokker-Planck equations. In the infinite-dimensional case, working with the field theory through functional Fokker-Planck equations such as \eqref{disc_functional_field_eqns} allows us to use tools with an essentially `differential equation' flavor such as spectral methods (\cite{rogers_demographic_2012, rogers_modes_2015}; See also Appendix \ref{App_examples} for a general model-independent pedagogical example), fast-mode (`adiabatic') elimination/slow manifold approximation~\citep{constable_stochastic_2013,parsons_dimension_2017}, and WKB style expansions to detect extreme events~(\cite{rogers_demographic_2012,assaf_wkb_2017}; though it bears noting that WKB theory is closely related to large deviation principles that occur in the study of Markov processes and often come in a distinctly `probabilistic' flavor, see~\cite{dembo_large_1998}). On the other hand, many biologically interesting questions are much easier to approach from the branching processes/SPDE side using tools that have an essentially `probabilistic' flavor such as duality~\citep{dawson_stochastic_1975,greenman_duality_2020}, infinitesimal generators~\citep{ethier_markov_1986,etheridge_mathematical_2011}, martingale theory~\citep{champagnat_unifying_2006,etheridge_mathematical_2011}, and weak/mild solutions to SPDEs~\citep{week_white_2021}. Rigorously establishing and systematically exploiting relations between functional Fokker-Planck equations of the form \eqref{disc_functional_field_eqns} and the sort of general SPDEs studied in~\cite{week_white_2021} for measure-valued branching processes could thus prove very fruitful for developing more integrative eco-evolutionary theory since it allows us to seamlessly transition between complementary views of the same object to suit the problem at hand\footnote{The physics literature has long recognized this. Physicists regularly exploit the connections between stochastic/quantum field equations and SPDEs, often studying SPDEs in completely `classical' systems using tools and language from quantum field theory~\citep{hochberg_effective_1999,baez_quantum_2018}.}.

@@

Currently, (stochastic) field equations are primarily used by physicists working in areas such as statistical field theory and quantum field theory, and are largely attacked using ingenious and often somewhat system-specific tools that may not necessarily be mathematically well understood~(\cite{carmona_stochastic_1999}; but see \cite{bogachev_fokker-planck-kolmogorov_2015} for a rigorous treatment of some infinite-dimensional Fokker-Planck equations). Equation \eqref{disc_functional_field_eqns} opens up the study of quantitative trait dynamics in finite fluctuating populations to analysis using some of these tools - For example, a careful reading of the literature provides many hints (albeit couched in physics language) for how (quantum) field theoretic ideas such as the Fock space representation~\citep{del_razo_probabilistic_2022}, `loop diagrams/expansions'~\citep{hochberg_effective_1999,dodd_many-body_2009}, and the often closely related path integral formalism~\citep{doi_second_1976, peliti_path_1985,hochberg_effective_1999, chow_path_2015, weber_master_2017}, can all be co-opted to coax biological insights from the sort of evolutionary dynamics modeled by functional equations of the form \eqref{disc_functional_field_eqns}.
